Return on investment of private sector malaria control at a large sugar facility in Southern Mozambique

### 2018-01-03 inst/rmd/03_results.Rmd
major update

---
title: Effectiveness of private sector malaria control at a large sugar
  facility in Southern Mozambique
date: ''
output: 
  pdf_document:
    includes:
      in_header: preamble.sty
fig_width: 4
fig_height: 2.6
bibliography: bibliography.bib
---

```{r, echo = FALSE, warning = FALSE, message = FALSE, comment = NA, error= FALSE, cache = FALSE}
# Packages
library(tidyverse)
library(knitr)
library(Hmisc)
library(brew)
library(maragra)
library(knitr)
library(RColorBrewer)
library(extrafont)
loadfonts()
## Global options
options(max.print="75")
opts_chunk$set(echo=FALSE,
	             cache=FALSE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE,
               dev = "cairo_pdf")
opts_knit$set(width=75)
options(xtable.comment = FALSE)

```{r}
source('paper_prepare.R')
```

\begin{center}
\begin{large}

```{r, results = 'asis', eval = TRUE}
make_authors()
```

\end{large}
\end{center}

\vspace{5mm}

\begin{center}
\textbf{Abstract}  
\end{center}

\vspace{5mm}
\begin{center}
\begin{changemargin}{3cm}{3cm}

\textbf{Context:} Incentivizing scaled-up private firms' involvement in malaria control and elimination activities requires a clear and robust demonstration of its benefits, and a quantification of which benefits can be expected to return to the firm. However, most studies to date examine benefits of malaria control purely from the perspective of the public sector. \textbf{Methods:} We analyze 4 years of malaria control and worker health and absenteeism data from a large sugar processing facility in Mozambique. We estimate the effect of indoor residual spraying (IRS) on worker health and absenteeism. \textbf{Results:} We demonstrate that the firm's engagement in malaria control activities is profitable, even if we only consider those benefits which directly affect the firm's bottom line and ignore secondary and societal benefits.

\end{changemargin}
\end{center}

# Introduction

Malaria has a nearly unquantifiably large economic impact on endemic societies. By affecting saving, investment [@Shretta2016], risk perception, productivity, absenteeism [@Nonvignon_2016], human capital accumulation [@CastelBranco2014], mortality, and costs of care [@Sachs2002], malaria likely has a negative effect on GDP and growth [@McCarthy_2000] [@Orem_2012]. Because of the relative affordability of most intereventions and the enormous societal costs of malaria, most forms of malaria control are cost-effective from the perspective of the government [@White_2011] [@Purdy_2013] [@Howard_2017].

From the persepctive of the private sector, however, investing in malaria control is not such a clear-cut case. Public health interventions targetting malaria - and their corresponding cost-effectiveness evaluations - most often pertain to the public sector. Accordingly, the analytical framework and metrics employed by these analyses most often focus on impact pertaining to the public good, such as an increase in life years adjusted for disability or quality  [@Goodman1999] [@Shretta2016] [@Lee2017] [@Hanson2004]. Though population-level health is certainly of importance to businesses, and improvements in health incidentally improve the economy at all levels [@World1999] [@Bloom2008] [@Vecchi_2013], these improvements may be too disperse or long-term to incentivize private sector involvement in health campaigns.

In the context of Mozambique, where a significant sector of the economy is dominated by a full large-scale foreing direct investment projects [@Robbins2012], the role of the private sector in health generally, and malaria specifically, is unequivocally important. Large agriculture and extractive industry firms take up wide swaths of land and employ hundreds of thousands  [@German2013]. The Mozambican state has encouraged large-scale entreprise with the aim of general economic development [@Buur2012]. And where large firms exist, they often take on social roles such as housing and health care [@Winkler]. At times, this role is necessary from a purely practical standpoint; in other cases, it is employed under the guise of "corporate social responsibility" [@Azemar2009]. Regardless of the language used, it is clear that private industry plays an important role in public health in Mozambique [@Robbins2012] [@CastelBranco2014].

Several cases exist of foreign firms engaging in large-scale malaria control campaigns [@Mouzin2011] [@Han] [@Bennett_2017] [@Kaula_2017]. But these studies generally consider population health as the outcome measure of interest, rather than worker absenteeism or productivity. Similarly, they often neglect to differentiate between those clinical costs which are absorbed by the local health system versus those which are absorbed by the firm itself. In the literature, making the "investment case" for malaria control or elimination generally implies that the investor is the public sector, and takes into account those costs and benefits which are applicable to the public sector [@Shretta_2017]; though appropriate in most cases (the public sector being the primary malaria control agent in most locations), the findings of these studies are rarely applicable to the private sector. In the case of a private firm not interested in "corporate social responsibility", it is not clear whether investing in malaria control would be profitable or not. This lack of certainty most likely discourages investment in many cases.

To address the question of the profitability of malaria control activities from the standpoint of a private firm, we analyze data during a 7 year period from a private sugar facility in Southern Mozambique. We assess the effect of indoor residual spraying (IRS) on the absenteeism and health of workers, and demonstrate that the firm's engagement in malaria control not only improved worker health, but also generated a positive return on investment.

# Methods

In collaboration with the sugar processing facility, Maragra Açucar SA (henceforth referred to as "Maragra"), we collected data for the period from January 2010 through December 2016. Data came from four sources: (i) the Human Resources' roster of worker details and absences, (ii) the facility's on-site clinic's medical and laboratory records, (iii) the facility's on-site malaria control program's records pertaining to the dates, chemicals, and location of IRS activities, and (iv) interviews with company employees pertaining to costs, data limitations, etc. Digitization and collection of data took place during the period from March 2016 through May 2017. Supplementary data pertaining to worker characteristics was obtained from through the Centro de Investigação em Saude de Manhiça's (CISM) demographic census, which covered workers from the district, but not those who migrated from other parts of the country [@Nhacolo_2006].

Data pertaining to district-wide malaria incidence was obtained from Mozambique’s Boletim Epidemiológico Semanal (BES), which is the system by which the National Malaria Control Program monitors incidence at the district level throghout the entire country, and reports
the number of confirmed weekly malaria cases at government health facilities. Using these case numbers, combined with population estimates from the National Statistical Institute (INE), we estimate each day's annualized weekly malaria incidence rate (cases per 1000 population at risk), interpolated from the weekly figures. We retrieved weather data for all Mozambican stations from NOAA. We estimated the meteorological conditions at the centroid of Manhiça using a simple interpolation method whereby the district's weather conditions were estimated to be a function of all Mozambican weather stations' reported conditions, inversely weighted by kilometers from district centroid.

Maragra regularly employs IRS at on-site worker households in order to reduce those workers' (and their families') risk of malaria infection. By using the CISM's census, we were able to identify from the record of houses sprayed (maintained by the company) who lived in those houses, as well as who lived off-site (and therefore did not receive IRS from Maragra). Then, using company HR and clinical records, we were able to identify absences and episodes of clinical malaria among all workers, as well as identify the time since the most recent IRS episode before the onsent of absence or illness.

Worker characteristics, illness and absenteeism data, along with IRS activity data, were systematically stored, collected, and used at the individual level by Maragra, and therefore of generally high quality. 
We sought to understand the effect of IRS on individual workers' likelihood of absence from work as well as their likelihood of clinical malaria. To estimate this effect, we estimated separate models for absence and illness. We employed interrupted time series [@Lopez_Bernal_2016] and a linear probability approach using the following econometric model.

$$
\hat{Y_{it}} = \beta_{0} +  (\beta_{1}) (\text{Season}_{t}) + (\beta_2{IRS}*\beta_3{IRS_t}) + ... + \epsilon
$$

$\hat{Y}$ is the rate of absence. $\beta_{1}$ represents the clinical malaria incidence at that time in the entire district of Manhiça.  Our demographic confounders (represented by $...$) are sex, age, and worker department (field, factory, or administrative). Our intervention was not a simple yes/no, but rather the product of whether the residence of the worker in question was treated in the last year, and, if so, the time since treatment (represented above as the interaction term, where where $_t$ represents time elapsed since commencement of the most recent IRS campaign). We define the malaria season as any time during which the clinical incidence of malaria in the district of Manhiça was at or greater than the median clinical incidence of malaria for the entire study period. These weeks are flagged as red in Figure 1, Panel A.  By using clinical incidence of the area of residence of the workers (as opposed to more typical proxies for malaria risk, such as only rainy vs. non rainy season), our seasonality estimate is a closer approximation of true malaria risk, incorporating lagged effects such as the incubation period of the parasite, as well as any inherent non-linear effects of weather. In addition, we adjust for daily precipitation

Because cost data was less systematically collected by Maragra, and because many costs could not be precisely quantified due to the abundance of in-kind and cross-departmental expenditures, we had to rely on rough estimations based on a mix of interviews, reciepts, and interpolations. Since our program cost data is not as reliable as our worker characteristic and outcome data, we were conservative in our estimates, and generally tried to err on the side of program activities and materials costing _more_ than what was reported, when doubt was aired. Cost data consisted of three types: (i) wages of malaria control employees, (ii) transporation and vehicle costs for IRS teams, and (iii) acquisition costs of purchasing IRS chemicals for fumigation (ACT and DDT), the latter two being combined into malaria control "programme" costs. We define the IRS program as "profitable" if return on investment is greater than 100%, ie if the savings associated with the estimated effect of IRS is greater than the costs of the program's administration.

Our formula for return on investment can be described in a straightforward fashion...

\begin{center}
$R = \dfrac{P_{w} - S_{wa} - S_{wc}}{P_{w}}$

\end{center}

...where $R$ is the return on investment, $P$ is the malaria control program's total operating cost, $w$ refers to costs at the per-worker level, $a$ refers to savings through avoided absences, and $ c $ refers to savings through avoided clinical encounters.

All data processing and analysis were carried out in R [@R] and all analysis code is freely available online [@brewgit]. Ethical approval for this project was obtained from the Institutional Ethics Review Board for Health at the CISM prior to data collection.

@@

```{r, results='asis'}
calculate <- function(var){
  out <- 
  model_data %>%
    mutate(year = format(date, '%Y')) %>%
    mutate(maragra_fabrica = ifelse(maragra_fabrica,
                                    'On site',
                                    'Off site'),
           rainy = ifelse(rainy,
                          'Rainy',
                          'Dry')) %>%
    filter(!is.na(rainy)) %>%
  group_by_(var, 'year') %>%
  summarise(absences = length(which(absent)),
            eligibles = length(absent),
            workers = length(unique(oracle_number))) %>%
  ungroup %>%
  mutate(absenteeism_rate = absences / eligibles * 100) 
  names(out)[1] <- 'variable'
  out$val <- paste0(round(out$absenteeism_rate, digits = 2),
                    '%')#,
                    # ')',
                    # '\n',
                    # ' (workers:', out$workers,
                    # ')')
  out <- 
    out %>%
    dplyr::select(variable, year, val)

  out <- out %>%
    spread(key = year,
           value = val)
  out$variable <- as.character(out$variable)
  new_var <- ifelse(var == 'maragra_fabrica',
                    'Residence',
                    ifelse(var == 'rainy',
                           'Precipitation',
                           var))
  out$variable[1] <- paste0(Hmisc::capitalize(new_var), ': ', out$variable[1])
  return(out)
}
out_list <- list()
vars <- c('season', 
           'department',
           'sex',
           'maragra_fabrica',
           'rainy')
for(i in 1:length(vars)){
  x <- calculate(var = vars[i])
  out_list[[i]] <- x
}
ab_table <- bind_rows(out_list)
names(ab_table)[1] <- ''
print(xtable::xtable(ab_table,
                     caption = 'Absenteeism rate by year and worker characteristics',
                     align = c('r' ,'r', rep('l', 4))),
      include.rownames = FALSE)
```

@@

```{r, fig.height = 3.5, fig.width = 6, fig.cap = 'Estimated effect of IRS on absenteeism by season'}
# Predict for visualization
dummy <- expand.grid(season = sort(unique(model_data$season)),
                     months_since = sort(unique(model_data$months_since)))
predictions <- predict(fit, dummy, se.fit = TRUE, interval = 'confidence')
dummy$predicted <- predictions$fit[,1]
dummy$lwr <- predictions$fit[,2]
dummy$upr <- predictions$fit[,3]
ggplot(data = dummy %>%
         filter(months_since != 'Never') %>%
         mutate(lwr = lwr * 100,
                upr = upr * 100,
                predicted = predicted * 100) %>%
         mutate(season = paste0(Hmisc::capitalize(as.character(season)), ' season')),
       aes(x = months_since,
           y = predicted,
           group = 1)) +
  geom_ribbon(aes(x = months_since,
                  ymin = lwr,
                  ymax = upr),
              fill = 'lightblue',
              alpha = 0.6) +
  # geom_line(color = 'blue', alpha = 0.5) +
  geom_point(color = 'blue', alpha = 0.5) + 
  geom_smooth(se = FALSE,
              color = 'black',
              # span = 10,
              alpha = 0.5) +
  facet_wrap(~season) +
  theme_maragra() +
  theme(axis.text.x = element_text(angle = 90, size = 8)) +
  labs(x = 'Months since IRS',
       y = 'Absenteeism (%)')
```

Details on full model here. Bla bla bla. See table 3.

```{r, results = 'asis'}
clean_up_model <- function(x){
  x = broom::tidy(x)
  for(j in 2:(ncol(x) -1)){
    x[,j] <- x[,j] * 100
  }
  x$term <- gsub('seasonhigh', 'Malaria season', x$term)
  x$term <- gsub('months_since', 'Months since IRS: ', x$term)
  x$term <- gsub('department', 'Department: ', x$term)
  x$term <- gsub('precipitation', 'Precipitation (mm) ', x$term)
  x$term <- gsub('maragra_fabricaTRUE', 'Living at factory', x$term)
  x$term <- gsub('sexM', 'Male', x$term)
  x$term <- gsub('permanent_or_temporaryTemporary', 'Temp contract', x$term)
  x$term <- gsub('rainyTRUE', 'Rainy day', x$term)
  names(x) <- Hmisc::capitalize(names(x))
  names(x) <- gsub('.', ' ', names(x), fixed = TRUE)
  x <- x %>% 
    mutate(Estimate = paste0(round(Estimate, 3), ' (P = ',
                             ifelse(`P value` <= 0.001,
                                    '< 0.001',
                                    round(`P value`, 4)), ')')) %>%
    dplyr::select(Term, Estimate)
  return(x)
}
# Combine them all
out_list <- list()
for (i in 1:length(models)){
  this_model <- letters[i]
  the_model <- models[[this_model]]
  the_model <- clean_up_model(the_model)
  names(the_model)[2] <- paste0('Model ', i)
  out_list[[i]] <- the_model
}
df <- out_list[[length(models)]]
for(i in (length(models) - 1):1){
  message(i)
    df <- left_join(df, out_list[[i]])
}
df <- df[,unique(c('Term', sort(names(df))))]
print(xtable::xtable(df, caption = 'Regression outputs'), size="\\fontsize{5pt}{6pt}\\selectfont")

@@

\vfill

\pagebreak

# Discussion

Some text goes here

# References

>>>>

@@

```{r, results='asis'}
calculate <- function(var){
  out <- 
  model_data %>%
    mutate(year = format(date, '%Y')) %>%
    mutate(on_site = ifelse(on_site,
                                    'On site',
                                    'Off site'),
           rainy = ifelse(rainy,
                          'Rainy',
                          'Dry')) %>%
    filter(!is.na(rainy)) %>%
  group_by_(var, 'year') %>%
  summarise(absences = length(which(absent)),
            eligibles = length(absent),
            workers = length(unique(oracle_number))) %>%
  ungroup %>%
  mutate(absenteeism_rate = absences / eligibles * 100) 
  names(out)[1] <- 'variable'
  out$val <- paste0(round(out$absenteeism_rate, digits = 1),
                    '%')#,
                    # ')',
                    # '\n',
                    # ' (workers:', out$workers,
                    # ')')
  out <- 
    out %>%
    dplyr::select(variable, year, val)

  out <- out %>%
    spread(key = year,
           value = val)
  out$variable <- as.character(out$variable)
  new_var <- ifelse(var == 'on_site',
                    'Residence',
                    ifelse(var == 'rainy',
                           'Precipitation',
                           var))
  left <- data.frame(x = c(new_var, rep('', (nrow(out) - 1))))
  out <- cbind(left, out)
  names(out)[1:2] <- c('Variable', 'x')
  # out$variable[1] <- paste0(Hmisc::capitalize(new_var), ': ', out$variable[1])
  return(out)
}
out_list <- list()
vars <- c('season', 
           'field',
          'permanent_or_temporary',
           'sex',
           'on_site',
           'rainy')
for(i in 1:length(vars)){
  x <- calculate(var = vars[i])
  out_list[[i]] <- x
}
ab_table <- bind_rows(out_list)
ab_table$Variable <- Hmisc::capitalize(ab_table$Variable)
ab_table$Variable[ab_table$Variable == 'Season'] <- 'Malaria season'
ab_table$Variable[ab_table$Variable == 'Field'] <- 'Worker type'
ab_table$Variable[ab_table$Variable == 'Permanent_or_temporary'] <- 'Contract'
ab_table$x <- Hmisc::capitalize(ab_table$x)
names(ab_table)[2] <- ''
print(xtable::xtable(ab_table,
                     caption = 'Absenteeism rate by year and worker characteristics',
                     align = c('r', 'r', 'l', rep('l', 4))),
      include.rownames = FALSE)
```

@@

```{r, fig.height = 3.5, fig.width = 6, fig.cap = 'Estimated effect of IRS on absenteeism by season'}
# Predict for visualization
dummy <- expand.grid(season = sort(unique(model_data$season)),
                     months_since = sort(unique(model_data$months_since)),
                     on_site = TRUE)
predictions <- predict(fit, dummy, se.fit = TRUE, interval = 'confidence', level = 0.99)
dummy$predicted <- predictions$fit[,1]
dummy$lwr <- predictions$fit[,2]
dummy$upr <- predictions$fit[,3]
ggplot(data = dummy %>%
         filter(months_since != 'Never') %>%
         mutate(lwr = lwr * 100,
                upr = upr * 100,
                predicted = predicted * 100) %>%
         mutate(season = paste0(Hmisc::capitalize(as.character(season)), ' season')),
       aes(x = months_since,
           y = predicted,
           group = 1)) +
  geom_ribbon(aes(x = months_since,
                  ymin = lwr,
                  ymax = upr),
              fill = 'lightblue',
              alpha = 0.6) +
  # geom_line(color = 'blue', alpha = 0.5) +
  geom_point(color = 'blue', alpha = 0.5) + 
  geom_line(color = 'blue', alpha = 0.5) +
  # geom_smooth(se = FALSE,
  #             color = 'black',
  #             # span = 10,
  #             alpha = 0.5) +
  facet_wrap(~season) +
  theme_maragra() +
  theme(axis.text.x = element_text(angle = 90, size = 8)) +
  labs(x = 'Months since IRS',
       y = 'Absenteeism (%)')
```

We create 4 worker fixed effects models. Different models for field vs not field, permanent vs temporary.

```{r, results = 'asis'}
clean_up_model <- function(x){
  # extract coefficients
  coefs <- data.frame(coef(summary(x)))
  # use normal distribution to approximate p-value
  coefs$p.z <- 2 * (1 - pnorm(abs(coefs$t.value)))
  ps <- coefs$p.z
  x = broom::tidy(x)
  for(j in 2:(ncol(x) -1)){
    x[,j] <- x[,j] * 100
  }
  x$term <- gsub('seasonhigh', 'Malaria season', x$term)    
  x$term <- gsub('seasonlow', 'Low malaria season', x$term)
  x$term <- gsub('months_since', 'Months since IRS: ', x$term)
  x$term <- gsub('department', 'Department: ', x$term)
  x$term <- gsub('precipitation', 'Precipitation (mm) ', x$term)
  x$term <- gsub('maragra_fabricaTRUE', 'Living at factory', x$term)
  x$term <- gsub('sexM', 'Male', x$term)
  x$term <- gsub('permanent_or_temporaryTemporary', 'Temp contract', x$term)
  x$term <- gsub('rainyTRUE', 'Rainy day', x$term)
  x$term <- gsub('on_siteTRUE', 'On site', x$term)
  x$term <- gsub('on_siteFALSE', 'Off site', x$term)
  x$term <- gsub('fieldNot field worker', 'Not field worker', x$term)
  names(x) <- Hmisc::capitalize(names(x))
  names(x) <- gsub('.', ' ', names(x), fixed = TRUE)
  x$`P value` <- NA
  x$`P value`[1:length(ps)] <- ps
  x <- x %>% 
    mutate(Estimate = paste0(round(Estimate, 3), ' (P',
                             ifelse(`P value` <= 0.001,
                                    '<0.001',
                                    paste0('=', round(`P value`, 3))
                                    ), ')')) %>%
    dplyr::select(Term, Estimate)
  x <- x %>% filter(!grepl('sd_', Term))
  return(x)
}

out_list <- list()
for(i in 1:length(fe_models)){
  message(i)
  this_model <- names(fe_models)[i]
  the_model <- fe_models[[which(names(fe_models) == this_model)]]
  the_model <- clean_up_model(the_model)
  names(the_model)[2] <- this_model
  out_list[[i]] <- the_model
}

# df <- out_list[[1]]
# for(i in 2:length(out_list)){
#   df <- left_join(df, out_list[[i]])
# }
# print(xtable::xtable(df, caption = 'Regression outputs'), size="\\fontsize{5pt}{6pt}\\selectfont",
#       include.rownames = FALSE)

df <- out_list[[1]]
namer <- names(df)[2]
names(df)[2] <- 'Estimate'

for(i in 2:length(out_list)){
  temp <- out_list[[i]]
  namer[i] <- names(temp)[2]
  names(temp)[2] <- 'Estimate'
  df <- bind_rows(df, temp)
}

library(kableExtra)
breaker <- nrow(out_list[[1]])
lengther <- length(out_list)
breaks <- c(0, breaker * 1:(lengther))
k <- kable(df, format = "latex", caption = "Models with worker fixed effects", booktabs = T) %>%
  kable_styling()
for (i in 1:(length(breaks)-1)){
  message(i)
  k <- k %>%
    group_rows(namer[i], 
               breaks[i] + 1,
               breaks[i] + breaker,
               latex_gap_space = "1.5em") 
}
print(k)

# # Combine them all
# out_list <- list()
# for (i in 1:length(models)){
#   this_model <- letters[i]
#   the_model <- models[[this_model]]
#   the_model <- clean_up_model(the_model)
#   names(the_model)[2] <- paste0('Model ', i)
#   out_list[[i]] <- the_model
# }
# df <- out_list[[length(models)]]
# # Order the Terms
# df_order <- c()
# for(i in 1:length(models)){
#   df_order <- c(df_order, out_list[[i]]$Term)
# }
# df_order <- unique(df_order)
# df <- left_join(data.frame(Term = df_order),
#                 df)
# for(i in (length(models) - 1):1){
#   message(i)
#     df <- left_join(df, out_list[[i]])
# }
# df <- df[,unique(c('Term', sort(names(df))))]
# print(xtable::xtable(df, caption = 'Regression outputs'), size="\\fontsize{5pt}{6pt}\\selectfont")

@@

## Return on investment

Details here on ROI calculation outcomes \todo{Not done yet. In conversation with Eduardo from MC about getting more accurate costs}

## Robustness and generalizability

Two principal concerns call into question the results of our analysis. First, the application of IRS to a workers house may be endogenous. To check for this  bla bla bla \todo{As per Menno's suggestion, will add details here pertaining to whether IRS predicts absenteeism before it is applied}. The second concern is that our quantification of consists is distorted by the fact that we treat IRS operations as essentially linear in nature, when in reality economies of scale, in-kind purchases and other factors likely make the true cost-per-spraying convex.\todo{Will address this with some sensitivity analysis}
